---
title: "Vaccination rate vs population-level factors"
author:
- affiliation: Baylor College of Medicine
  name: Ashish Damania
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    toc: true
    toc_depth: 2
---

<style type="text/css">
.main-container {
  max-width: 1800px;
  margin-left: auto;
  margin-right: auto;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width = 600)
```


```{r}
library(tidyverse)
```
 
# Read vax location data
```{r, message=FALSE, results='hide', warning=FALSE}
file_list_vax_loc <- list.files(path="./Vax_location_data",pattern='*.csv', full.names = T)
df_list_vax_loc <- lapply( file_list_vax_loc, read_csv)


df_vax_loc <- do.call(rbind, df_list_vax_loc)
```


# Read vax type data
```{r, message=FALSE, results='hide', warning=FALSE}
file_list_vax_type <- list.files(path="./Vax_type_data/",pattern='*.csv', full.names = T)
df_list_vax_type <- lapply(file_list_vax_type, read_csv)


df_vax_type <- do.call(rbind, df_list_vax_type)
```


# Combine vax type and location data    

- one VAERS_ID may have multiple VAX_TYPE since advserse incident can happen for multiple vaccines - which is reflected by the multiple VAERS_ID in the VAX_TYPE data.  

- We are restricting the date to MMR vaccine and age of less or equal to five.  

```{r}
df_vax_combined <- df_vax_type %>%
  inner_join(df_vax_loc, by = c("VAERS_ID")) %>%
  filter(VAX_TYPE=="MMR") %>%
  filter(!is.na(STATE)) %>% # Remove rows with no location information
 #uses state.name variable which has all us states variable names as a list
  mutate(STATE_FULL_NAME=state.name[match(STATE,state.abb)]) %>% 
  mutate(YEAR = substring(RECVDATE,7,10)) %>%
  filter(AGE_YRS <= 5) 
```


## Using dummy states dataframe to fill zeroes in states with no adverse reaction reported
```{r}
dummy_df_states <- data.frame(rep(state.name,23),seq(1995,2017)) %>%
                  rename("state"=1, "year"=2) %>%
                  mutate(state=as.character(state)) %>%
                  mutate(year=as.numeric(year))

df_vax_zero_fixed <- df_vax_combined %>%
  filter(!is.na(STATE_FULL_NAME)) %>%
  mutate(YEAR=as.numeric(YEAR)) %>%
  select(HOSPITAL, STATE_FULL_NAME,YEAR) %>% #Get Hospitalization column
  count(STATE_FULL_NAME, YEAR, HOSPITAL) %>% 
  spread(HOSPITAL,n) %>% # Count hospitalization  
  filter(YEAR < 2018) %>%
  right_join(dummy_df_states, by =c("STATE_FULL_NAME"="state",YEAR="year"))
```






# Get Population data to calculate rate of vaccine adverse reaction
```{r}
df_population_age_year <- 
  read_tsv("Bridged-Race Population Estimates 1990-2018.txt") %>%
  filter(`Age Group` =="1-4 years") %>%
  filter(!is.na(`Yearly July 1st Estimates`)) %>%
  filter(State!="District of Columbia") %>%
  filter(`Yearly July 1st Estimates` > 1994 & `Yearly July 1st Estimates` < 2018) %>%
  select(`Age Group`, State, `Yearly July 1st Estimates`,`Population`)
```





#Get Educational attainment data

## 1995-2000
```{r}
list_education_list_95_2000 <- 
  list.files(path="Education_attainment",pattern='*.xlsx', full.names = T)

df_95_2000<- data_frame(filename=list_education_list_95_2000) %>%
  mutate(file_contents =map(list_education_list_95_2000, ~ readxl::read_xlsx(file.path(.) ))) %>%
  mutate(filename = str_remove(filename,'(.)*_')) %>%
  mutate(filename = str_remove(filename,'.xlsx')) %>%
  select(year=filename,file_contents) 

df_95 <- df_95_2000 %>%
  filter(year==1995) %>%
  unnest() %>%
  slice(2:n()) %>%
  rename("state_name"=2) %>%
  select(year, state_name,percent_bachelors_or_higher=`...5`) %>%
     mutate(state_name= gsub('[^ -~]', '', state_name)) %>%
     mutate(state_name = iconv(state_name, "UTF-8", "ASCII", sub="")) %>%
     mutate(state_name = trimws(state_name,which=c("both"))) %>%
     filter(state_name %in% state.name) 


df_96 <- df_95_2000 %>%
  filter(year==1996) %>%
  unnest() %>%
  slice(2:n()) %>%
  rename("state_name"=2) %>%
  select(year, state_name,percent_bachelors_or_higher=`...5`) %>%
     mutate(state_name= gsub('[^ -~]', '', state_name)) %>%
     mutate(state_name = iconv(state_name, "UTF-8", "ASCII", sub="")) %>%
     mutate(state_name = trimws(state_name,which=c("both"))) %>%
     filter(state_name %in% state.name) 


df_97 <- df_95_2000 %>%
  filter(year==1997) %>%
  unnest() %>%
  slice(2:n()) %>%
  rename("state_name"=2) %>%
  select(year, state_name,percent_bachelors_or_higher=`...5`) %>%
     mutate(state_name= gsub('[^ -~]', '', state_name)) %>%
     mutate(state_name = iconv(state_name, "UTF-8", "ASCII", sub="")) %>%
     mutate(state_name = trimws(state_name,which=c("both"))) %>%
     filter(state_name %in% state.name) 


df_98 <- df_95_2000 %>%
  filter(year==1998) %>%
  unnest() %>%
  slice(2:n()) %>%
  rename("state_name"=2) %>%
  select(year, state_name,percent_bachelors_or_higher=`...5`) %>%
     mutate(state_name= gsub('[^ -~]', '', state_name)) %>%
     mutate(state_name = iconv(state_name, "UTF-8", "ASCII", sub="")) %>%
     mutate(state_name = trimws(state_name,which=c("both"))) %>%
     filter(state_name %in% state.name) 


df_99 <- df_95_2000 %>%
  filter(year==1999) %>%
  unnest() %>%
  slice(2:n()) %>%
  rename("state_name"=2) %>%
  select(year, state_name,percent_bachelors_or_higher=`...5`) %>%
     mutate(state_name= gsub('[^ -~]', '', state_name)) %>%
     mutate(state_name = iconv(state_name, "UTF-8", "ASCII", sub="")) %>%
     mutate(state_name = trimws(state_name,which=c("both"))) %>%
     filter(state_name %in% state.name) 
  

df_2000 <- df_95_2000 %>%
  filter(year==2000) %>%
  unnest() %>%
  slice(2:n()) %>%
  rename("state_name"=2) %>%
  select(year, state_name,percent_bachelors_or_higher=`...4`) %>%
     mutate(state_name= gsub('[^ -~]', '', state_name)) %>%
     mutate(state_name = iconv(state_name, "UTF-8", "ASCII", sub="")) %>%
     mutate(state_name = trimws(state_name,which=c("both"))) %>%
     filter(state_name %in% state.name) 


df_95_2000_com <- rbind(df_95,df_96,df_97,df_98,df_99,df_2000)
```





## 2001-2006
```{r}
list_education_list_2001_2006 <- list.files(path="Education_attainment",pattern='*.xls$', full.names = T)
```

```{r}
df_2001_2002 <- data_frame(filename=list_education_list_2001_2006) %>%
  mutate(file_contents =map(filename, ~ readxl::read_xls(file.path(.)))) %>%
  mutate(filename = str_remove(filename,'(.)*_')) %>%
  mutate(filename = str_remove(filename,'.xls')) %>%
  filter(filename==c(2001,2002))%>%
  unnest() %>%
  select(filename,2,`...5` ) %>% #5th column is the percentage of the population of 25 and above
  janitor::clean_names() %>%
  slice(6:n()) %>%
  rename(state_name=`table_with_row_headers_in_column_a_and_column_headers_in_row_5_and_repeated_in_row_41`) %>%
  filter(state_name %in% state.name) %>%
  rename(percent_bachelors_or_higher=x5)

df_2003_2004 <- data_frame(filename=list_education_list_2001_2006) %>%
  mutate(file_contents =map(filename, ~ readxl::read_xls(file.path(.)))) %>%
  mutate(filename = str_remove(filename,'(.)*_')) %>%
  mutate(filename = str_remove(filename,'.xls')) %>%
  filter(filename%in% c(2003:2004,1)) %>%
  unnest() %>%
  select(filename,2,`...5` ) %>% #5th column is the percentage of the population of 25 and above
  janitor::clean_names() %>%
  slice(5:n()) %>%
  rename(state_name=`table_with_row_headers_in_column_a_and_column_headers_in_row_5`) %>%
  filter(state_name %in% state.name) %>%
  rename(percent_bachelors_or_higher=x5)


df_2005_2006 <- data_frame(filename=list_education_list_2001_2006) %>%
  mutate(file_contents =map(filename, ~ readxl::read_xls(file.path(.)))) %>%
  mutate(filename = str_remove(filename,'(.)*_')) %>%
  mutate(filename = str_remove(filename,'.xls')) %>%
  filter(filename%in% c(2005:2006,1)) %>%
  unnest() %>%
  select(filename,2,`...5` ) %>% #5th column is the percentage of the population of 25 and above
  janitor::clean_names() %>%
  slice(5:n()) %>%
  rename(state_name=`table_with_row_headers_in_column_a_and_column_headers_in_rows_5_and_6`) %>%
  filter(state_name %in% state.name) %>%
  rename(percent_bachelors_or_higher=x5)

df_2001_2006_com <- rbind(df_2001_2002,df_2003_2004,df_2005_2006) %>%
                     rename(year=filename)
  
```





## 2007-2017
```{r}
list_education_list_2007_2017 <- list.files(path="Education_attainment",pattern='*.csv', full.names = T)



df1_2007_2017 <- data_frame(filename=list_education_list_2007_2017) %>%
  mutate(file_contents =map(filename, ~ read_csv(file.path(.)))) %>%
  mutate(filename = str_remove(filename,'.*/')) %>%
  mutate(filename = str_remove(filename,'_S1501_with_ann.csv')) %>%
    mutate(filename = case_when(
    filename == "ACS_07_3YR" ~ 2007,
    filename == "ACS_08_3YR" ~ 2008,
    filename == "ACS_09_5YR" ~ 2009,
    filename == "ACS_10_5YR" ~ 2010,
    filename == "ACS_11_5YR" ~ 2011,
    filename == "ACS_12_5YR" ~ 2012,
    filename == "ACS_13_5YR" ~ 2013,
    filename == "ACS_14_5YR" ~ 2014,
    filename == "ACS_15_5YR" ~ 2015,
    filename == "ACS_16_5YR" ~ 2016,
    filename == "ACS_17_5YR" ~ 2017
  )) 


df1_2007_2009 <- df1_2007_2017 %>%
  filter(filename %in% 2007:2009) %>%
  unnest() %>%
  select(filename,`GEO.display-label`,89) %>%
  select(year=filename,state_name = `GEO.display-label`,percent_bachelors_or_higher=`HC01_EST_VC15`) %>%
  slice(2:n()) %>%
  filter(state_name %in% state.name) 

df1_2010_2014 <- df1_2007_2017 %>%
  filter(filename %in% 2010:2014) %>%
  unnest() %>%
 select(filename,`GEO.display-label`,89) %>%
  select(year=filename,state_name = `GEO.display-label`,percent_bachelors_or_higher=`HC01_EST_VC17`) %>%
  slice(2:n()) %>%
  filter(state_name %in% state.name) 

df1_2015_2017 <- df1_2007_2017 %>%
  filter(filename %in% 2015:2017) %>%
  unnest() %>%
   select(year=filename,state_name=`GEO.display-label`,percent_bachelors_or_higher=HC02_EST_VC18) %>%
  slice(2:n()) %>%
  filter(state_name %in% state.name) 
  


df1_2007_2017_com <- rbind(df1_2007_2009, df1_2010_2014, df1_2015_2017)

```

### Combine all dataframes to get educational attainment
```{r}
df_educational_attainment <- rbind(df_95_2000_com,df_2001_2006_com,df1_2007_2017_com) %>%
  mutate_at(vars(year,percent_bachelors_or_higher),as.numeric)
```





# Join Population data with Combined vax type and location data
```{r}
df_pop_reaction_rate <- df_vax_zero_fixed %>%
  rename(adverse_reaction_count=4, hospitalization_count=3) %>%
  mutate(hospitalization_count= replace_na(hospitalization_count,0))%>%
   mutate(adverse_reaction_count= replace_na(adverse_reaction_count,0))%>%
  inner_join(df_population_age_year, by=c("YEAR"="Yearly July 1st Estimates","STATE_FULL_NAME"="State")) %>%
  inner_join(df_educational_attainment, by=c("YEAR"="year","STATE_FULL_NAME"="state_name")) %>%
  mutate(reaction_rate_per_100_000 = (adverse_reaction_count/Population)*100000) %>%
  mutate(hospitalization_rate_per_100_000 = (hospitalization_count/Population)*100000) %>%
  select(STATE_FULL_NAME, YEAR, reaction_rate_per_100_000, hospitalization_rate_per_100_000,percent_bachelors_or_higher) 
  
```







# Read Vaccination rate data from 1995-2017  

- This spreadsheet includes data for multiple vaccines for ages 13 months - 35 months  

```{r}
#Declare years as character to select 
years <- as.character(c(1995:2017))

df_vax_rate <- readxl::read_excel("dataView2116_17.xls", skip = 2, range = "A3:FA142") %>% #confirm that the range for the 18-35 months for MMR vaccination is as specified here
  select(Names,!!years) %>%
  filter(Names %in% state.name) %>% #state.name is a builtin variable with all US states name
  gather(year,rate,-Names) %>%
  mutate(year=as.numeric(year), rate = as.numeric(rate))
```


# Combine vax rate with vax adverse rate
```{r}
df_final <- df_pop_reaction_rate %>%
  inner_join(df_vax_rate, by=c("STATE_FULL_NAME"="Names", "YEAR"="year"))

```










# Restrict Analysis to state that allows voluntary philosophical vaccine exemptions  

- Using sources from these references:  
  - Omer, S. B., Pan, W. K. Y., Halsey, N. A., Stokley, S., Moulton, L. H., Navar, A. M., …    Salmon, D. A. (2006). Nonmedical exemptions to school immunization requirements: secular trends and association of state policies with pertussis incidence. JAMA: The Journal of the American Medical Association, 296(14), 1757–1763.  


  - Omer, S. B., Richards, J. L., Ward, M., & Bednarczyk, R. A. (2012). Vaccination policies and rates of exemption from immunization, 2005-2011. The New England Journal of Medicine, 367(12), 1170–1171.  
  

https://www.cdc.gov/vaccines/imz-managers/coverage/schoolvaxview/pubs-resources.html#pubs

```{r}
states_philosophical_exemptions_1995_2004 <- c("California", "Colorado", 
                                               "Idaho", "Louisiana", "Maine", 
                                               "Michigan", "Minnesota","New Mexico", 
                                               "North Dakota", "Ohio", "Oklahoma", 
                                               "Oregon", "Pennsylvania", "Utah",
                                               "Vermont","Washington", "Wisconsin")

states_philosophical_exemptions_2005_2011 <-  c("Arizona","Arkansas","California", 
                                                "Colorado", "Idaho", "Louisiana",
                                                "Maine", "Michigan", "Minnesota",
                                                "New Mexico", "North Dakota", "Ohio",
                                                "Oklahoma", "Oregon", "Pennsylvania",
                                                "Texas","Utah","Vermont",
                                                "Washington", "Wisconsin")

states_philosophical_exemptions_2012_2015 <-  c("Arizona","Arkansas","California", 
                                                "Colorado", "Idaho", "Louisiana", 
                                                "Maine", "Michigan", "Minnesota", 
                                                "North Dakota", "Ohio", "Oklahoma",
                                                "Oregon", "Pennsylvania","Texas",
                                                "Utah","Vermont","Washington", "Wisconsin")

states_philosophical_exemptions_2016_2017 <-  c("Arizona","Arkansas", "Colorado", 
                                                "Idaho", "Louisiana", "Maine", 
                                                "Michigan","Minnesota", "North Dakota", 
                                                "Ohio", "Oklahoma","Oregon",
                                                "Pennsylvania","Texas","Utah","Washington", 
                                                "Wisconsin")
```


**California had philosophical based exemption until 2015**  

**Vermont had atleast until 2004**  

**Arizona has no philosophical based exemption until 2004**




## Add these states to the dataframe
```{r}
df_f2 <- df_final %>%
  mutate(exemption_y_n = case_when(
    STATE_FULL_NAME %in% states_philosophical_exemptions_1995_2004 & (YEAR >= 1995 & YEAR <= 2004) ~ 1,
    STATE_FULL_NAME %in% states_philosophical_exemptions_2005_2011 & (YEAR >= 2005 & YEAR<= 2011) ~ 1,
    STATE_FULL_NAME %in% states_philosophical_exemptions_2012_2015 & (YEAR >= 2012 & YEAR <=2015 ) ~ 1,
    STATE_FULL_NAME %in% states_philosophical_exemptions_2016_2017 & (YEAR >= 2016 & YEAR <= 2017) ~ 1,
    TRUE ~ 0
  ))  %>%
  mutate(exemption_y_n = factor(exemption_y_n)) 
``` 

# Save data as R object file
```{r}
saveRDS(df_f2,"processed_intermediate_data.rds")
```



# Summary of the dataset
```{r}
Hmisc::describe(df_f2)
summarytools::descr(df_f2)
```


# Data dictionary
```{r}
Variables <- colnames(df_f2)
data_type <- c("text", 
               "Numeric", 
               "Numeric", 
               "Numeric", 
               "Numeric",
               "Numeric",
               "text")
description <- c("US state name",
                 "Year",
                 "Adverse reaction rate per 100,000 using VAERS database and CDC wonder database of population from 1-4 ages",
                 "Hospitalization rate per 100,000 using VAERS database and CDC wonder database of population from 1-4 ages",
                 "Percent of population among 25 or higher with bachelors degree or higher",
                 "MMR vaccination percent coverage in the State",
                 "Indication of whether or not if a State allows non-medical personal exemption. `0` indicates that State does not have non-medical vaccine exemption. `1` indicates that State allows non-medical personal exmeption for vaccination")


tibble(Variables,data_type,description) %>%
knitr::kable()
```

# Table 1
```{r}
reaction_rate_df <- df_f2 %>%
  group_by(exemption_y_n) %>%
do(data.frame(rbind(Hmisc::smean.cl.boot(.$reaction_rate_per_100_000)))) %>%
  ungroup() %>%
  mutate(factor = paste0("Reaction_rate_per_100_000"))


hospitalization_rate_df <-df_f2 %>%
  group_by(exemption_y_n) %>%
do(data.frame(rbind(Hmisc::smean.cl.boot(.$hospitalization_rate_per_100_000)))) %>%
  ungroup() %>%
  mutate(factor = paste0("Hospitalization_rate_per_100_000"))


percent_bachelors_or_higher_df  <- df_f2 %>%
  group_by(exemption_y_n) %>%
do(data.frame(rbind(Hmisc::smean.cl.boot(.$percent_bachelors_or_higher)))) %>%
  ungroup() %>%
  mutate(factor = paste0("Percent Bachelors or higher"))

vaccination_rate_df <- df_f2 %>%
  group_by(exemption_y_n) %>%
do(data.frame(rbind(Hmisc::smean.cl.boot(.$rate)))) %>%
  ungroup() %>%
   mutate(factor = paste0("Vaccination Rate"))


rbind(reaction_rate_df,
      hospitalization_rate_df,
      percent_bachelors_or_higher_df,
      vaccination_rate_df) %>%
      select(-Upper,-Lower) %>%
      spread(exemption_y_n,Mean)

```





```{r}
df_f2 %>%
  select(reaction_rate_per_100_000, hospitalization_rate_per_100_000, percent_bachelors_or_higher, rate) %>%
  gather(key,value) %>%
  group_by(key) %>%
  do(data.frame(rbind(Hmisc::smean.cl.boot(.$value))))
  
```


```{r}
df_f2 %>%
  group_by(exemption_y_n) %>%
  summarytools::descr(.)
```














# Spearman Correlation analysis
```{r}
df_f3 <- df_f2 %>%
  mutate(exemption_y_n = as.numeric(as.character(exemption_y_n)) )

PerformanceAnalytics::chart.Correlation(df_f3[,-1], method = c("spearman"))
```




# Linear cross-random effects univariate

## Reaction rate only
```{r}
lmer_reaction_only <- lmerTest::lmer(rate ~ reaction_rate_per_100_000 + (1|YEAR) + (1|STATE_FULL_NAME) ,data=df_f2, REML = F)
summary(lmer_reaction_only)
```



## Hospitalization rate only
```{r}
lmer_hospitalization_only <- lmerTest::lmer(rate ~ hospitalization_rate_per_100_000 +
                                              (1|STATE_FULL_NAME) + (1|YEAR),
                                            data=df_f2, REML = F)
summary(lmer_hospitalization_only)
```


## Education attainment only
```{r}
lmer_percent_bachelors_or_higher <- lmerTest::lmer(rate ~ percent_bachelors_or_higher + 
                                                     (1|YEAR)+ (1|STATE_FULL_NAME),
                                                   data=df_f2, REML = F)
summary(lmer_percent_bachelors_or_higher)
```


## Philosophical based exemption
```{r}
lmer_exemption_y_n <- lmerTest::lmer(rate ~ 
                                       exemption_y_n + 
                                       (1|YEAR) + 
                                       (1|STATE_FULL_NAME),
                                     data=df_f2, REML = F)
summary(lmer_exemption_y_n)
```



# Test the variability between the State vs the years
```{r}
lmer0 <- lmerTest::lmer(rate ~ 
                          1 + 
                          (1|STATE_FULL_NAME) + 
                          (1|YEAR), 
                        data=df_f2, REML =F)
summary(lmer0)
```


# Cross-random effects model (random intercept model) for state and year as "Random" and other covariates as fixed  

- We are using this model since YEAR and States are **NOT** nested within each other.  

```{r}
lmer1 <- lmerTest::lmer(rate ~  
                          reaction_rate_per_100_000 + 
                          hospitalization_rate_per_100_000 + 
                          exemption_y_n + 
                          percent_bachelors_or_higher +
                          (1|YEAR) +
                          (1|STATE_FULL_NAME),
                          data=df_f2, REML=F)
summary(lmer1)
```

-  This model shows that none of the independent variables or predictors are significant at p-value of 0.05



# Logistic regression model with coverage greater than 93% vs less than 93%

```{r}
df_f4 <- df_f2 %>%
  mutate(rate_90_y_n = case_when(
    rate >= 93 ~ 1,
    rate < 93 ~ 0
  ) ) %>%
  filter(YEAR > 1994)
```

## Univariate


### Reaction rate per 100,000
```{r}
m4_1 <- lme4::glmer(rate_90_y_n ~ 
                    reaction_rate_per_100_000 +
                          (1|YEAR) +
                          (1|STATE_FULL_NAME),
                          data=df_f4, family="binomial")
summary(m4_1)

exp(m4_1@beta)
```

### Hospitalization rate per 100,000
```{r}
m4_2 <- lme4::glmer(rate_90_y_n ~ 
                    hospitalization_rate_per_100_000 +
                          (1|YEAR) +
                          (1|STATE_FULL_NAME),
                          data=df_f4, family="binomial")
summary(m4_2)

exp(m4_2@beta)
```


### Education attainment
```{r}
m4_3 <- lme4::glmer(rate_90_y_n ~ 
                    percent_bachelors_or_higher   +
                          (1|YEAR) +
                          (1|STATE_FULL_NAME),
                          data=df_f4, family="binomial")
summary(m4_3)

exp(m4_3@beta)
```


### Exemption policy
```{r}
m4_4 <- lme4::glmer(rate_90_y_n ~ 
                    exemption_y_n   +
                          (1|YEAR) +
                          (1|STATE_FULL_NAME),
                          data=df_f4, family="binomial")
summary(m4_4)

exp(m4_4@beta)
```



## Multivariate
```{r}
m4 <- lme4::glmer(rate_90_y_n ~ 
                          hospitalization_rate_per_100_000 +
                          percent_bachelors_or_higher  +
                          (1|YEAR) +
                          (1|STATE_FULL_NAME)+
                          exemption_y_n,
                          data=df_f4, family="binomial")
(m4_summary <- summary(m4))
broom::tidy(m4_summary$coefficients)
```

- Here the model tells whether or not the coverage was greater or less than 93% relates with the predictors.  
- 93% percent or more is required for herd immunity

https://www.nejm.org/doi/full/10.1056/NEJMp1905099 (93-95%)
https://jamanetwork.com/journals/jamapediatrics/fullarticle/2203906 (95%)
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6626669/ (92-95%)







## Odds ratio with confidence interval
```{r}
odds_ratio_logistic <-lme4::confint.merMod(m4,method = c("Wald"))

(odds_ratio_conf_int_df <- broom::tidy(odds_ratio_logistic) %>%
  janitor::clean_names() %>%
  mutate(x2_5=exp(x2_5),x97_5=exp(x97_5)) %>%
  slice(3:n()) %>%
  cbind(broom::tidy(exp(m4@beta))) %>%
    filter(rownames!="(Intercept)"))
```



### Plot odds ratio with confidence interval
```{r}

odds_ratio_conf_int_df %>%
  mutate(rownames=case_when(
    rownames=="hospitalization_rate_per_100_000" ~ "Hospitalization rate per 100,000",
    rownames=="percent_bachelors_or_higher" ~"Percent bachelors or higher",
    rownames=="exemption_y_n1" ~ "Personal belief exemption policy"
  )) %>%
  mutate(color = if_else(x > 1, "Positive", "Negative")) %>%
ggplot(data=., aes(x=reorder(rownames,x),y=x,color=color)) +
  geom_pointrange(aes(ymin = x2_5, ymax = x97_5),size=1.5) +
  scale_color_manual(
    values = c("Positive" = "blue", "Negative" = "red"),
    name = ""
  ) +
  xlab("Factors") +
  ylab("Odds Ratio") +
  guides(fill = guide_legend(title = "LEFT", title.position = "left")) +
  coord_flip()+
  geom_hline(yintercept = 1,linetype = 2)+
  theme_bw()+
  theme(text = element_text(size=18))

 ggsave("Logistic_regression_odds_ratio.pdf", width = 25, height = 15, units = "cm")
```













# Check for multicollinearity for all perdictors
## Check for Variance inflation factor

### Linear model
```{r}
car::vif( lmerTest::lmer(rate ~ reaction_rate_per_100_000 + 
                               hospitalization_rate_per_100_000 +
                               exemption_y_n +
                               percent_bachelors_or_higher + (1|YEAR) + (1|STATE_FULL_NAME),
                               data=df_f2, REML =F) )


```

**VIF values for all predictors are below 3 which indicates low correlation between the variables.**  


### Logistic model
```{r}
 car::vif(lme4::glmer(rate_90_y_n ~ 
                    reaction_rate_per_100_000 +
                          hospitalization_rate_per_100_000 +
                          percent_bachelors_or_higher  +
                          (1|YEAR) +
                          (1|STATE_FULL_NAME)+
                          exemption_y_n,
                          data=df_f4, family="binomial"))
```





















