---
title: "Create plots and figures for the project"
description: |
  Impact of different factors on Vaccination Rate
author:
  - name: Ashish Damania as part of the project 
    affiliation: Baylor College of Medicine
date: "`r Sys.Date()`"
output:
   radix::radix_article:
     toc: true
     self_contained: true
---

```{r}
library(tidyverse)
library(rmarkdown)
library(broom)
library(tableHTML)
library(geofacet)
library(gghighlight)
library(sf)
library(gridExtra)
devtools::install_github("hrbrmstr/albersusa")
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


# Data for model
```{r}
complete_data <- readRDS("processsed_data/processed_intermediate_data.rds") 
paged_table(complete_data)
```





# Rate of vaccination across the years 1995-2017
```{r}
ggplot(complete_data) +
  geom_line(aes(YEAR, rate,color = STATE_FULL_NAME)) +
   geom_hline(yintercept=93,linetype="dashed") +
  facet_geo(~ STATE_FULL_NAME, grid = "us_state_grid2") +
  theme_bw()+
  
  scale_x_continuous(labels = function(x) paste0("'", substr(x, 3, 4))) +
  ylab("Vaccination Rate (%)")+
  theme(legend.position = "none") +
  ggtitle("Vaccination rate across 50 US states 1995-2017")

ggsave("figures/rate_of_vaccination_states_1995_2017.pdf", width = 30, height = 20, units = "cm")
```


```{r}
ggplot(complete_data) +
  geom_line(aes(YEAR, rate,color = STATE_FULL_NAME)) +
  gghighlight(max(rate) > 95,use_direct_label = FALSE)+
  facet_geo(~ STATE_FULL_NAME, grid = "us_state_grid2") +
  theme_bw()+
  
  scale_x_continuous(labels = function(x) paste0("'", substr(x, 3, 4))) +
  ylab("Vaccination Rate (%)")+
  theme(legend.position = "none") +
  ggtitle("States which have achieved 95% vaccination rate atleast once since 1995")

ggsave("figures/rate_of_vaccination_states.pdf", width = 30, height = 20, units = "cm")
```




# Rate of vaccination across the years 2010-2017
```{r}
complete_data %>%
  filter(YEAR > 2009) %>%
ggplot(.) +
  geom_line(aes(YEAR, rate,color = STATE_FULL_NAME)) +
  gghighlight(max(rate) > 95,use_direct_label = FALSE)+
  facet_geo(~ STATE_FULL_NAME, grid = "us_state_grid2") +
  theme_bw()+
  
  scale_x_continuous(labels = function(x) paste0("'", substr(x, 3, 4))) +
  ylab("Vaccination Rate (%)")+
  theme(legend.position = "none") +
  ggtitle("States which have achieved 95% vaccination rate atleast once since 2010")

ggsave("figures/rate_of_vaccination_states_2010_2017.pdf", width = 30, height = 20, units = "cm")
```


# Rate of adverse reaction rate across the years 1995-2017
```{r}
complete_data %>%
ggplot(.) +
  geom_line(aes(YEAR, reaction_rate_per_100_000)) +
  #gghighlight(max(rate) > 95,use_direct_label = FALSE)+
  facet_geo(~ STATE_FULL_NAME, grid = "us_state_grid2") +
  theme_bw()+
  
  scale_x_continuous(labels = function(x) paste0("'", substr(x, 3, 4))) +
  ylab("Reaction rate per 100,000")+
  theme(legend.position = "none") +
  ggtitle("Reaction rate per 100,000")

ggsave("figures/rate_of_adverse_reaction_states_1995_2017.pdf", width = 30, height = 20, units = "cm")
```




# Rate of hospitalization rate across the years 1995-2017
```{r}
complete_data %>%
ggplot(.) +
  geom_line(aes(YEAR, hospitalization_rate_per_100_000)) +
  #gghighlight(max(rate) > 95,use_direct_label = FALSE)+
  facet_geo(~ STATE_FULL_NAME, grid = "us_state_grid2") +
  theme_bw()+
  
  scale_x_continuous(labels = function(x) paste0("'", substr(x, 3, 4))) +
  ylab("Hospitalization rate per 100,000")+
  theme(legend.position = "none") +
  ggtitle("Hospitalization rate per 100,000")

ggsave("figures/rate_of_hospitalization_rate_states_1995_2017.pdf", width = 30, height = 20, units = "cm")
```





## States that allow personal exemption
```{r}
df1_plotmeans <- complete_data %>%
  select(STATE_FULL_NAME,rate,YEAR,exemption_y_n) %>%
  filter (YEAR > 1994) %>%
  group_by(STATE_FULL_NAME,exemption_y_n) %>%
do(data.frame(rbind(Hmisc::smean.cl.boot(.$rate)))) %>%
  ungroup() %>%
  filter(exemption_y_n == 1)


  p2_yes_policy <- ggplot(data=df1_plotmeans, aes(x=reorder(STATE_FULL_NAME,-Mean),y=Mean)) +
  geom_pointrange(aes(ymin = Lower, ymax = Upper,color=exemption_y_n),size=1) + 
    scale_color_manual(name="",labels=c("","") ,
    values=c("red","blue") )+
      xlab("States") +
    ylab("Mean vaccination coverage 1995-2017") +
  coord_flip()+
  theme_bw()+
    theme(text = element_text(size=18))+
    theme(legend.position = "none")+
  annotate("rect", ymin = 93, ymax = Inf, xmin=Inf,xmax = -Inf,alpha = .2,fill="green")
   ggsave("figures/Mean_vaccination_coverage_1995_2017_yes_exemption.pdf", width = 30, height = 20, units = "cm")
```  
  
  
## States that do no allow personal exemption  
```{r}  
df3_plotmeans <- complete_data %>%
  select(STATE_FULL_NAME,rate,YEAR,exemption_y_n) %>%
  filter (YEAR > 1994) %>%
  group_by(STATE_FULL_NAME,exemption_y_n) %>%
do(data.frame(rbind(Hmisc::smean.cl.boot(.$rate)))) %>%
  ungroup() %>%
  filter(exemption_y_n == 0) %>%
  filter(!is.na(Lower))


 p1_no_policy <-  ggplot(data=df3_plotmeans, aes(x=reorder(STATE_FULL_NAME,-Mean),y=Mean)) +
  geom_pointrange(aes(ymin = Lower, ymax = Upper,color=exemption_y_n),size=1) + 
  scale_color_manual(name="",labels=c("Does not allow Philosophical exemption") ,
  values=c("blue") )+
  xlab("States") +
  ylab("Mean vaccination coverage 1995-2017") +
  guides(fill = guide_legend(title = "LEFT", title.position = "left")) +
  coord_flip()+
  theme_bw()+
  theme(text = element_text(size=18))+
  theme(legend.position = "none")+
  annotate("rect", ymin = 93, ymax = Inf, xmin=Inf,xmax = -Inf,alpha = .2,fill="green")
```


```{r}
df2_plotmeans <- complete_data %>%
  select(STATE_FULL_NAME,rate,YEAR,exemption_y_n) %>%
  filter (YEAR > 1994) %>%
  mutate(exemption_y_n= case_when(
    exemption_y_n =="0" ~ "Does not allow philisophical exemption",
    exemption_y_n =="1" ~ "Allows philisophical exemption"
  )) %>%
  group_by(STATE_FULL_NAME,exemption_y_n) %>%
do(data.frame(rbind(Hmisc::smean.cl.boot(.$rate)))) %>%
  ungroup()%>%
  arrange(-Mean) %>%
  mutate(r=row_number())
 # filter(exemption_y_n == 0)


ggplot(data=df2_plotmeans, aes(x=reorder(STATE_FULL_NAME,-Mean),y=Mean)) +
  geom_pointrange(aes(ymin = Lower, ymax = Upper,color=exemption_y_n)) +
  scale_color_manual(name="State",labels=c("Allows Philosophical Exemption","Does not allow Philosophical exemption") ,
                     values=c("red","blue") )+
  xlab("States") +
  ylab("Mean vaccination coverage 1995-2017") +
  guides(fill = guide_legend(title = "LEFT", title.position = "left")) +
  coord_flip()+
  theme_bw()+
  annotate("rect", ymin = 90, ymax = 100, xmin=Inf,xmax = -Inf,alpha = .2,fill="green")

```

## Combination graph of mean coverage States with no PBE (Blue) vs States with PBE (Red)
```{r}
p1_2 <- gridExtra::grid.arrange(p1_no_policy, p2_yes_policy, nrow = 1)
ggsave("figures/combined_mean_coverage.pdf",p1_2,width = 30, height = 20, units = "cm")
```





# Current coverage
```{r}
com_df_2017 <- complete_data %>%
  filter(YEAR==2017) %>%
  select(STATE_FULL_NAME,rate) %>%
  filter(STATE_FULL_NAME!="Alaska") %>%
  filter(STATE_FULL_NAME!="Hawaii") %>%
  mutate(category = case_when(
    rate >= 93 ~ "At herd immunity",
    rate > 90  & rate  < 93 ~ "Above 90% and below 93%",
    rate <=90 & rate > 85 ~ "Below 90%"
  ))

all_df <- states %>%
  inner_join(com_df_2017, by = c("NAME"="STATE_FULL_NAME"))

ggplot() +
  geom_sf(data=all_df, aes(fill=factor(category)))+
 scale_fill_manual(values=c( "#F3DFA2","#7EBDC2","#BB4430","#962389"), 
                       breaks=c( "At herd immunity","Above 90% and below 93%","Below 90%"),
                       labels=c("At herd immunity","Above 90% and below 93%","Below 90%")) +
  geom_label(data = nc3_coords, aes(X, Y, label = NAME), size = 2.5, fontface = "bold", 
        nudge_y = nc3_coords$nudge_y, nudge_x = nc3_coords$nudge_x) +
  theme_bw()+
    guides(fill=guide_legend(title="Vaccination coverage (2017)")) +
    theme(legend.position="top")+
   theme(axis.text.x=element_blank(), 
        axis.text.y=element_blank(),
        axis.ticks.x=element_blank(),
        axis.ticks.y=element_blank()) +
  xlab("") +
  ylab("")
```









# Heatmap for the count of years States meeting or exceeding herd immunity
```{r}
com_df_2 <- complete_data %>%
  mutate(herd_immunity=case_when(
    rate >= 93 ~ 1,
    rate < 93 ~ 0
  )) %>%
  group_by(STATE_FULL_NAME,herd_immunity) %>%
  tally() %>%
  spread(herd_immunity,n) %>%
  mutate(.,`1`=ifelse(is.na(`1`), 0, `1`)) %>%
  mutate(percent_achieved_herd_immunity = `1`/23) %>%
  rename("count_of_achieving_herd_immunity"=`1`)


states <- albersusa::usa_sf() %>%
  inner_join(com_df_2, by = c("name"="STATE_FULL_NAME")) %>%
   mutate(Herd_immunity_count=case_when(
        `count_of_achieving_herd_immunity` >= 0 & `count_of_achieving_herd_immunity` <= 5 ~ "0 to 5",
        `count_of_achieving_herd_immunity` >=6 & `count_of_achieving_herd_immunity` <=10 ~ "6 to 10",
        `count_of_achieving_herd_immunity` >=11 & `count_of_achieving_herd_immunity` <=15 ~ "11 to 15",
        `count_of_achieving_herd_immunity` >=16 & `count_of_achieving_herd_immunity` <=23 ~ "16 to 23",
      ))
  

nc3_points2 <- sf::st_point_on_surface(states)
nc3_coords2 <- as.data.frame(sf::st_coordinates(nc3_points2))
nc3_coords2$NAME <- states$iso_3166_2

    
ggplot() +
  geom_sf(data=states, aes(fill=as.factor(Herd_immunity_count)))+
  geom_label(data = nc3_coords2, aes(X, Y, label = NAME), size = 4, fontface = "bold", 
        nudge_y = nc3_coords2$nudge_y, nudge_x = nc3_coords2$nudge_x) +
   scale_fill_manual(values=c("#fec4b0", "#c03e38","#720206","#f78170","#ea5349","#fffff"), 
                        name="",
                        breaks=c("0 to 5","6 to 10", "11 to 15", "16 to 23"),
                        labels=c("0 to 5","6 to 10", "11 to 15", "16 to 23"))+
  theme_bw()+
    guides(fill=guide_legend(title="Count of years")) +
    theme(legend.position="bottom")+
   theme(axis.text.x=element_blank(), 
        axis.text.y=element_blank(),
        axis.ticks.x=element_blank(),
        axis.ticks.y=element_blank(),
        panel.grid = element_blank(),
        legend.position = c(0.9, 0.3)
      ) +
  xlab("") +
  ylab("")
 ggsave("figures/Heatmap2_count_years_exceeding_93_percent.pdf", width = 30, height = 20, units = "cm")
```






























